---
title: "Homework 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,eval=T,message=F,warning=F,fig.align='center')
library(tidyverse)
```

## 1. Independence of Events <small>(8 points, 2 points each)</small>

A fair four-sided die (numbered 1,2,3 and 4) is rolled three times. Consider the following three events:

A: "The sum is 6 or less"

B: "The second roll is more than the first"

C: "The sum is odd"

D: "Exactly two of the rolls are the same"

```{r}
set.seed(123)
rolls <- replicate(10000, sample(1:4, size = 3, replace = TRUE))
event_A <- colSums(rolls) <= 6
event_B <- rolls[2, ] > rolls[1, ]
event_C <- colSums(rolls) %% 2 == 1
event_D <- apply(rolls, 2, function(x) length(unique(x)) == 2)
independence_checks <- list(
  A_B = mean(event_A & event_B) == mean(event_A) * mean(event_B),
  C_D = mean(event_C & event_D) == mean(event_C) * mean(event_D),
  A_C = mean(event_A & event_C) == mean(event_A) * mean(event_C),
  B_D = mean(event_B & event_D) == mean(event_B) * mean(event_D)
)
independence_checks
```

a. Are A and B independent?

> No, A and B are not independent.

b. Are C and D independent?

> No, C and D are not independent.

c. Are A and C independent?

> No, A and C are not independent.

d. Are B and D independent?

> No, B and D are not independent.

## 2. Stock Portfolio <small>(8 points, 2 points each)</small>

Consider two stocks. Stock A has a mean annual return of 0.05 and a standard deviation of 0.01. Stock B has a mean annual return of 0.10 and a standard deviation of 0.04. Past data suggests that the prices have a correlation of -0.65.
You have \$1000 to invest in a portfolio composed of these two stocks. 

a. Suppose you invest \$500 in stock A and \$500 in stock B. What will be the expected return and the standard deviation of your portfolio (in dollars)? 
*Hint:* $SD(aX+bY) = \sqrt{Var(aX+bY)}= \sqrt{a^2Var(X)+2abCov(X,Y)+b^2Var(Y)}$ and $Cov(X,Y)=\sigma_X \sigma_Y \rho_{XY}$

```{r}
mean_A <- 0.05  
sd_A <- 0.01      
mean_B <- 0.10   
sd_B <- 0.04    
cor_AB <- -0.65    
cov_AB <- sd_A * sd_B * cor_AB
total_investment <- 1000
weight_A <- 0.5  
weight_B <- 0.5  

expected_return_percent <- weight_A * mean_A + weight_B * mean_B
expected_return_dollars <- total_investment * expected_return_percent
expected_return_dollars 
portfolio_sd_percent <- sqrt(weight_A^2 * sd_A^2 + weight_B^2 * sd_B^2 + 2 * weight_A * weight_B * cov_AB)
portfolio_sd_dollars <- total_investment * portfolio_sd_percent
portfolio_sd_dollars 
```

b. Give an expression for the standard deviation for a portfolio investing $\$x$ in stock A and $\$1000-x$ in stock B. 

```{r}
portfolio_sd_x <- function(x) {
  weight_A <- x / 1000        
  weight_B <- (1000 - x) / 1000
  sqrt(weight_A^2 * sd_A^2 + weight_B^2 * sd_B^2 + 2 * weight_A * weight_B * cov_AB)
}
```

c. Find the value $x$ that minimizes this standard deviation. 
```{r}
optimal_x <- optimize(portfolio_sd_x, c(0, 1000))$minimum
optimal_x 
```

d. What is the expected return and standard deviation (in dollars) for the portfolio split you chose in part c?
```{r}
optimal_weight_A <- optimal_x / 1000
optimal_weight_B <- (1000 - optimal_x) / 1000
optimal_expected_return_percent <- optimal_weight_A * mean_A + optimal_weight_B * mean_B

optimal_expected_return_dollars <- total_investment * optimal_expected_return_percent
optimal_expected_return_dollars

optimal_portfolio_sd_dollars <- total_investment * portfolio_sd_x(optimal_x)
optimal_portfolio_sd_dollars
```

## 3. Mousetrap: A Rube Goldberg Machine  <small>(5 points, 1 points each)</small>

Consider the board game Mousetrap. When you trigger the trap, the ball needs to 

1. roll down a **ramp**,
2. be carried up by a **hammer**,
3. roll  down another **ramp**,
4. go down a bathtub **drain**,
5. land on a **lever** which
6. sends a diver into a **basin** which then
7. causes the **trap** to descend.

Suppose you can model the success of each component as independent Bernoulli random variables $X_1$ through $X_7$, where the probability of $X_i$ succeeding is $.94$ for each component. 

a. What is the probability that the trap works correctly from beginning to end?

```{r}
p_success <- 0.94
p_all_success <- p_success^7
p_all_success
```

b. If you attempt the trap 15 times during a game, how many times do you expect it to work correctly from beginning to end?

```{r}
n_attempts <- 15
expected_successes <- n_attempts * p_all_success
expected_successes
```

c. Suppose you let the marble go: Let $Y$ be the number of components of the trap that work in a row. If one component fails, then you stop counting. Calculate the probability mass function of $Y$. *Hints: $Y$ can have values $0, 1, 2, 3, 4, 5, 6$ or $7$. This is like a Geometric r.v. except it does not have infinite support. How can you use this fact to determine the pmf? Be careful: what is a 'success' in this context?*

```{r}
pmf_Y <- function(k) {
  if (k < 7) {
    return((0.94^k) * (1 - 0.94))
  } else {
    return(0.94^7)
  }
}

pmf_values <- sapply(0:7, pmf_Y)
pmf_values
```

d. Calculate the expected value of $Y$ and its variance. 

```{r}
expected_Y <- sum(0:7 * pmf_values)
expected_Y
expected_Y2 <- sum((0:7)^2 * pmf_values)
variance_Y <- expected_Y2 - expected_Y^2
variance_Y
```

e. What is $P(Y = 6 | Y \geq 3)$?

```{r}
p_Y_ge_3 <- sum(pmf_values[4:8])
p_Y_6_given_ge_3 <- pmf_values[7] / p_Y_ge_3
p_Y_6_given_ge_3
```

## 4. Bayes Rule with forecasting <small>(9 points, 3 points each)</small>

Madison, Wisconsin gets about 115 days of precipitation per year (we'll just call it rain for the remainder of this problem, even though it is often snow). On days when it ends up raining the forecast on Channel 3000 is correct (i.e. the morning forecast calls for rain) 93% of the time. On days where there is no rain the morning forecast correctly says there will be no rain 87% of the time. 

a. Calculate the overall probability that a forecast on a randomly selected day is correct.

```{r}
P_rain <- 115 / 365           
P_no_rain <- 1 - P_rain  
P_forecast_rain_given_rain <- 0.93 
P_forecast_no_rain_given_no_rain <- 0.87 

P_correct <- P_forecast_rain_given_rain * P_rain + P_forecast_no_rain_given_no_rain * P_no_rain
P_correct
```

b. If the weather from day to day is independent and the forecast accuracy from day to day is independent, then how many correctly forecasted days do you expect in a 365 day year? 

```{r}
expected_correct_days <- 365 * P_correct
expected_correct_days
```

c. If the forecast calls for rain, what is the likelihood that it actually rains? In other words, find $P(\text{Rain} | \text{Forecast Rain})$.

```{r}
P_forecast_rain <- P_forecast_rain_given_rain * P_rain + (1 - P_forecast_no_rain_given_no_rain) * P_no_rain
P_rain_given_forecast_rain <- (P_forecast_rain_given_rain * P_rain) / P_forecast_rain
P_rain_given_forecast_rain
```
