---
title: "Homework 11"
author: "Ria Sharma"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,eval=T,message=F,warning=F,fig.align='center')
library(tidyverse)
```

## Problem 1. Warm up: Log-Odds, Odds and Probability <small>5 points; 1 point each</small>

### Part a) 
Convert a log-odds of 1.25 to probability.

```{r}
log = 1.25
prob = 1/(1 + exp(-log))
prob
```

***

0.7772999

***

### Part b) 
Convert probability of 0.65 to odds.

```{r}
prob = 0.65
odds = prob/(1-prob)
odds
```

***

1.857143

***

### Part c)
Event $A$ has a probability of 0.8 and the odds ratio of $A$ to $B$ is 1.3. Calculate the probability of event $B$.

```{r}
p_a = 0.8
o_a_b = 1.3
o_a = p_a/(1-p_a)
o_b = o_a/o_a_b
p_b = o_b/(1 + o_b)
p_b
```

***

0.754717

***

### Part d) 
You roll a single 6 sided die. What are the odds you get a number divisible by 3?

```{r}
p_divisible_by_3 = 2 / 6
o_divisible_by_3 = p_divisible_by_3 / (1 - p_divisible_by_3)
o_divisible_by_3
```

***

0.5

***

### Part e) 
The odds ratio comparing event A to B is 9 and the risk ratio is 3. What is $Pr(A)$? *Hint: let $Pr(A)=a$ and $Pr(B)=b$, and write out the odds ratio and risk ratio as equations. 

***

- **Risk Ratio (RR):**
\[
RR = \frac{P(A)}{P(B)} = 3 \quad \Rightarrow \quad P(A) = 3P(B)
\]

- **Odds Ratio (OR):**
\[
OR = \frac{\frac{P(A)}{1 - P(A)}}{\frac{P(B)}{1 - P(B)}} = 9 \quad \Rightarrow \quad \frac{P(A)(1 - P(B))}{P(B)(1 - P(A))} = 9
\]

### Step-by-Step Solution:

1. Substitute \( P(A) = 3P(B) \) into the Odds Ratio equation:
\[
\frac{3(1 - P(B))}{1 - 3P(B)} = 9
\]

2. Solve for \( P(B) \):
\[
3(1 - P(B)) = 9(1 - 3P(B)) \quad \Rightarrow \quad P(B) = \frac{1}{4}
\]

3. Use \( P(A) = 3P(B) \) to find:
\[
P(A) = 3 \cdot \frac{1}{4} = \frac{3}{4}
\]

### Final Verification:

- **Risk Ratio:**
\[
RR = \frac{P(A)}{P(B)} = \frac{\frac{3}{4}}{\frac{1}{4}} = 3 \quad \text{(Correct)}
\]

- **Odds Ratio:**
\[
OR = \frac{\frac{P(A)}{1 - P(A)}}{\frac{P(B)}{1 - P(B)}} = \frac{\frac{\frac{3}{4}}{1 - \frac{3}{4}}}{\frac{\frac{1}{4}}{1 - \frac{1}{4}}}
= \frac{\frac{\frac{3}{4}}{\frac{1}{4}}}{\frac{\frac{1}{4}}{\frac{3}{4}}}
= 9 \quad \text{(Correct)}
\]

### Conclusion:
\[
P(A) = \frac{3}{4}, \quad P(B) = \frac{1}{4}
\]

***


## Problem 2. Interpreting logistic regression <small>6pts; 2 pts each</small>

Suppose we collect data for a group of students in a statistics class with independent variables $X_{1}=\text{hours studied}$, $X_{2}=\text{GPA}$, and binary response variable
$$
Y= \begin{cases} 1 &\mbox{ if student received an A} \\
  0 &\mbox{ otherwise. }
  \end{cases}
$$
Suppose that we fit a logistic regression model to the data, predicting $Y$ from $X_1$ and $X_2$ (and an intercept term) and produce estimated coefficients $\hat{\beta}_{0}=-5, \hat{\beta}_{1}=0.06, \hat{\beta}_{2}=1.1$.

### Part a) Logistic regression and probability
According to our fitted model, what is the probability that a student receives an A if they study for $40$ hours and have a GPA of $3.5$?

```{r}
z <- -5 + (0.06 * 40) + (1.1 * 3.5)
prob <- 1 / (1 + exp(-z))
prob
```

### Part b) Interpreting coefficients
According to our fitted model, an additional hour spent studying is associated with *how much* of an increase in the log odds of receiving an A?

```{r}
beta_1 <- 0.06
log_odds_increase <- beta_1
log_odds_increase
```

> According to the fitted model, an additional hour spent studying is associated with an increase of 0.06 in the log-odds of receiving an A, holding GPA constant.

### Part c) "Inverting" logistic regression probabilities
According to our fitted model, how many hours would the student in Part (a) need to study to have a $50\%$ chance of getting an A in the class?
That is, keeping GPA fixed at $3.5$, how many hours of study are needed so that the predicted probability of an A is $50\%$?
If you aren't up for the math, feel free to find an approximate solution via guess-and-check in R.

```{r}
beta_0 <- -5
beta_1 <- 0.06
beta_2 <- 1.1
gpa <- 3.5

probability <- function(hours) {
  z <- beta_0 + beta_1 * hours + beta_2 * gpa
  return(1 / (1 + exp(-z)))
}

hours <- 19.17  
prob <- probability(hours)
prob
```

> The student would need to study ~19.17 hours to have a 50% chance at an A.

## Problem 3. Palmer Penguins Part 1  <small>9pts; 3 pts each</small>

The Palmer Penguin dataset (https://allisonhorst.github.io/palmerpenguins/) consists of 344 observations of penguins belonging to 3 penguin species across the islands in the Palmer Archipeligo in Antarctica. We will build a logistic model attempting to classify the penguins based on physical characteristics. For each penguin we've recorded: 

* `species` - the species, either "Gentoo", "Adelie" or "Chinstrap"
* `island` - which of three islands the Penguin was observed (Biscoe, Dream or Torgersen)
* `bill_length_mm` - the length of the bill in mm
* `bill_depth_mm` - the depth of the bill (vertical thickness of the closed bill)
* `filler_length_mm` - length of their cute flippers
* `body_mass_g` - the body mass in grams
* `sex` - female, male or NA (unknown)
* `year` - The year of the observation: 2007, 2008 or 2009

First you need to download the library. Run this chunk of code once.
```{r, eval=FALSE, echo=FALSE}
#Run this code once to install the library
install.packages("palmerpenguins")
```

Then load the library and the penguin dataset. Note: Your RMD won't knit until you run the above chunk.
```{r}
library(palmerpenguins)
```

### a) Adelie penguins based on island

We are going to try to classify penguins as **Adelie** or **not Adelie**. So create a new variable called `Adelie` which will be 1 or 0 based on whether the penguin species is Adelie

Perform some analysis looking at each of the 3 islands - create a 2 way table between `island` and the `Adelie` variable. Look at the proportions conditioned on island. What proportion of observations on each island were Adelie?

```{r}
penguins$Adelie <- ifelse(penguins$species == "Adelie", 1, 0)

adelie_table <- table(penguins$island, penguins$Adelie)
adelie_table

adelie_proportions <- prop.table(adelie_table, margin = 1)
adelie_proportions
```

> On Biscoe Island, 26% of the penguins were Adelie On Dream Island, 45% of the penguins were Adelie On Torgersen Island, 100% of the penguins were Adelie

### b) Adelie Penguins on Dream

Find the (i) probability, (ii) odds and (iii) log-odds that a randomly selected penguin from Dream is an Adelie penguin?

```{r}
adelie_dream <- 56
total_dream <- 124

#probability
probability <- adelie_dream / total_dream

#odds
odds <- probability / (1 - probability)

#log-odds
log_odds <- log(odds)
```

```{r}
probability  
odds         
log_odds
```

### c) An island-based classifier

Now fit a logistic model predicting whether a penguin is Adelie based on island.
Interpret the intercept and the coefficient of the `islandDream` variable. Use this model to predict the probability that a penguin from Dream is Adelie.


Logistic Model:
```{r}
adelie_model <- glm(Adelie ~ island, data = penguins, family = "binomial")
summary(adelie_model)
```

> Interpret Intercept: The intercept (−1.0361) represents the log-odds of a penguin being an Adelie if the penguin is from the reference island (Biscoe). Penguins on Biscoe have about a 26.1% probability of being Adelie.

> Interpret islandDream Coefficent: The coefficient for islandDream (0.8419) represents the change in log-odds of being an Adelie penguin for penguins on Dream Island compared to Biscoe. Penguins on Dream are approximately 2.32 times more likely to be Adelie than those on Biscoe.

Predicted Probability:
```{r}
z_dream <- coef(adelie_model)["(Intercept)"] + coef(adelie_model)["islandDream"]
prob_dream <- 1 / (1 + exp(-z_dream))
prob_dream
```

> Penguins on Dream have about a 45.2% probability of being Adelie.

## Problem 4. Penguins Part II <small>10pts; 2 pts each</small>

In this problem we will work once again with the Palmer Penguin dataset. We will work with a subset by taking out all missing values. After you have installed the package and loaded the library, uncomment the line below.
```{r}
 penguins.complete <- penguins[complete.cases(penguins),]
```

### a) Predicting Palmer Penguins with quantitative Predictors

Now use the two bill measurements (`bill_length_mm` and `bill_depth_mm`) as predictors in a new logistic model. Suppose a penguin with a bill length of 54.1 and a bill depth of 23.7 is observed. What is the model's probability that the penguin is an Adelie penguin? *Hint: Your response variable in the formula can be `species=="Adelie"` and you won't have to create a new binary response variable column.*

```{r}
adelie_model <- glm(species == "Adelie" ~ bill_length_mm + bill_depth_mm,
                    data = penguins.complete,
                    family = "binomial")
summary(adelie_model)
```

```{r}
coef(adelie_model)
bill_length <- 54.1
bill_depth <- 23.7

z <- coef(adelie_model)["(Intercept)"] +
     coef(adelie_model)["bill_length_mm"] * bill_length +
     coef(adelie_model)["bill_depth_mm"] * bill_depth

prob_adelie <- 1 / (1 + exp(-z))
prob_adelie
```

> The predicted probability that a penguin with a bill length of 54.1 mm and a bill depth of 23.7 mm is an Adelie is approximately 33.88%.

### b) Interpreting coefficients

Are longer bills associated with an increased or decreased likelihood that a penguin is an Adelie penguin?

> From the logistic regression model, the coefficient for bill_length_mm is −2.2099, indicating that longer bills are associated with a decreased likelihood of being an Adelie penguin. For every 1 mm increase in bill length, the log-odds of being an Adelie decreases by 2.2099, holding bill depth constant.

### c) A full classifier

Fit a logistic model to predict whether a penguin is a **Chinstrap** penguin using all four of the biological measurements (`bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`). 


Which of the predictors are significant?

```{r}
chinstrap_model <- glm(species == "Chinstrap" ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g,
                       data = penguins.complete,
                       family = "binomial")

summary(chinstrap_model)
```

- **Significant Predictors** (\( p < 0.05 \)):
  - `bill_length_mm`: (\( p = 0.00113 \)) - A strong positive association with being a Chinstrap penguin.
  - `body_mass_g`: (\( p = 0.00129 \)) - A strong negative association with being a Chinstrap penguin.

- **Non-Significant Predictors**:
  - `bill_depth_mm`: (\( p = 0.78114 \)) - Not significant.
  - `flipper_length_mm`: (\( p = 0.06220 \)) - Marginally significant, showing a weak negative association.

> From the model, bill length and body mass are the most reliable biological measurements for predicting whether a penguin is a Chinstrap penguin. While flipper length shows a marginal relationship, bill depth does not significantly contribute to the prediction.

### d) Assessing the model
Suppose you will predict that a penguin is a Chinstrap if the estimated $\hat{y}=\sigma(\hat{z})>0.5$. When predicting using this threshold, what is the type 1 error rate of your predictor on the dataset? What is power of the predictor on the dataset? 
*Hint: you will want to compare the predicted $\hat{y}$ values to the actual $y$ values. The `table` command can produce a 2x2 confusion matrix to help you answer this question.*

```{r}
chinstrap_probs <- predict(chinstrap_model, type = "response")
chinstrap_preds <- ifelse(chinstrap_probs > 0.5, 1, 0)
chinstrap_actual <- as.numeric(penguins.complete$species == "Chinstrap")
confusion_matrix <- table(Predicted = chinstrap_preds, Actual = chinstrap_actual)
print(confusion_matrix)

type1_error_rate <- confusion_matrix[2, 1] / sum(confusion_matrix[, 1])
power <- confusion_matrix[2, 2] / sum(confusion_matrix[, 2])
type1_error_rate

power
```

#### Metrics:
- **Type I Error Rate**:
  \[
  \text{Type I Error Rate} = \frac{\text{False Positives}}{\text{Total Non-Chinstrap Observations}} = \frac{1}{265} \approx 0.0038 \, (0.38\%).
  \]

- **Power**:
  \[
  \text{Power} = \frac{\text{True Positives}}{\text{Total Chinstrap Observations}} = \frac{65}{68} \approx 0.9559 \, (95.59\%).
  \]

> These results indicate that the model has a very low Type I error rate and high power, making it effective at identifying Chinstrap penguins while minimizing false positives.

### e) Adjusting the Type 1 error rate

Now modify your threshold from 0.5 to some other threshold with the goal of achieving the highest power possible while keeping the type 1 error rate  below 0.05. What threshold would you use? What is the type 1 error rate and power of this new classifier?

```{r}
thresholds <- seq(0.1, 0.9, by = 0.05)
penguin_df <- penguins.complete
model <- glm(species == "Chinstrap" ~ bill_length_mm + bill_depth_mm + 
               flipper_length_mm + body_mass_g, 
             data = penguin_df, 
             family = "binomial")
max_power <- 0
best_threshold <- NA
type_1_error_at_best <- NA

for (i in seq_along(thresholds)) {
  penguin_df$predicted_prob <- predict(model, type = "response")
  penguin_df$predicted_class <- ifelse(penguin_df$predicted_prob > thresholds[i], 1, 0)
  confusion_matrix <- table(penguin_df$predicted_class, penguin_df$species == "Chinstrap")
  type1_error <- confusion_matrix[2, 1] / (confusion_matrix[1, 1] + confusion_matrix[2, 1])
  power <- confusion_matrix[2, 2] / (confusion_matrix[1, 2] + confusion_matrix[2, 2])
  
  if (type1_error < 0.05 && power > max_power) {
    max_power <- power
    best_threshold <- thresholds[i]
    type_1_error_at_best <- type1_error
  }
}

paste("Max power we can get is", max_power, 
      "at the threshold of", best_threshold, 
      "with a Type 1 error rate of", round(type_1_error_at_best, 3))
```

> Max Power: The model achieves perfect power (1.0).

> Best Threshold: 0.1 is the optimal threshold.

> Type 1 Error Rate: 3.4%, which is below the 5% limit.
