---
title: "FA24 STAT340 Midterm <small>take-home portion</small>"
output: html_document
---

<style>h2{margin-top:5rem}h3,h4:not([class]){margin-top:4rem}
.main-container{padding:2rem}</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,eval=T,warning=F,message=F,
                      fig.width=5,fig.height=4,fig.align="center")
```


## Q1 - Testing

Infinity Cable (fictional) manufactures fiber optic cable for industrial purposes. They have three manufacturing facilities in Alderwood, Brentwick and Coppergrove. A recent quality control audit sampled various 1 kilometer lengths of cable from the three facilities and counted the number of flaws in the cables:

```{r}
cables = data.frame(
  site = rep(c("Alderwood","Brentwick","Coppergrove"), times=c(53,48,51)),
  flaws = c(6,4,4,1,2,1,4,2,4,3,4,8,2,4,6,2,4,1,2,2,0,2,5,2,3,3,3,2,5,4,4,
            1,4,3,5,4,4,3,3,4,0,3,4,4,3,5,3,2,1,1,2,3,4,2,5,1,2,2,3,1,2,3,
            0,4,2,4,2,2,2,4,4,2,3,5,2,3,2,2,3,1,3,1,1,1,1,0,3,4,4,4,2,2,4,
            3,3,2,1,7,3,1,1,3,6,3,7,4,2,3,1,0,4,1,3,4,8,3,3,1,4,3,3,2,2,3,
            3,1,0,4,6,3,3,3,7,3,4,3,2,2,4,3,1,4,1,5,3,3,2,3,3,1,3,1)
)
```

Your job is to perform a comparison using the testing techniques we've covered in this course to see if there is evidence that the facilities actually produce cable of different quality. Do not use techniques we have not covered in this course.

a. (2pts) A colleague suggests that we should focus our attention on a comparison of Alderwood and Brentwick (ignoring Coppergrove), since they seem to have the largest gap in average number of flaws. Why specifically would this not be a good approach?

> We should not focus soley on Alderwood and Brentwick while we ignore Coppergrove becuase we are then not considering all the available data and by including Coppergrove we may have critical insights into the overall quality differences accross all facilities. If we only pick the largest gap in average number of flaws, it can lead to bias results and increase the likelihood of type | errors. Overall, this can be harmful to drawing complete, acurate, and contextual conclusions based on the full dataset.

b. Now you will perform the test. You should

   i. (1pt) State your null and alternative hypotheses, 
   ii. (1pt) decide on the test statistic to use, 
   iii. (1pt) state whether this is a one or two-sided test, and explain why, 
   iv. (2pts) produce a simulated distribution for your test statistic under the null hypothesis 
   v. (2pts) calculate the $p$-value
   vi. (1pt) make your conclusion

*Hint: your test statistic should somehow compare all 3 groups. There are many possible statistics you could use. If you are stuck consider a statistic based on the site means or the site variances.*

> ***PLEASE WRITE YOUR WORK BELOW:***

> Null Hypothesis (H0): There is no difference in the mean number of flaws among the three facilities (Alderwood, Brentwick, Coppergrove).

> Alternative Hypothesis (HA): At least one facility has a different mean number of flaws compared to the others.

> Test Statistic: A statistic based on the variance between group means and the variance within groups: \[
F = \frac{\text{Between-Group Variance}}{\text{Within-Group Variance}}
\]

> Two Sided Test: We are testing whether there are any differences among the group means without specifying which group(s) might differ.

> Simulated Null Distribution: The null distribution was generated by shuffling the flaws across groups (assuming the null of no differences) and recalculating the test statistic for each shuffle.


```{r}
# insert code here
#calc test stat
group_means <- tapply(cables$flaws, cables$site, mean)
group_sizes <- table(cables$site)
overall_mean <- mean(cables$flaws)

#group var
between_group_var <- sum(group_sizes * (group_means - overall_mean)^2) / (length(group_means) - 1)
within_group_var <- sum(tapply(cables$flaws, cables$site, function(x) sum((x - mean(x))^2))) / 
                    (nrow(cables) - length(group_means))

#test stat
observed_stat <- between_group_var / within_group_var

#simulate null distribution
set.seed(42)
num_simulations <- 10000
null_stats <- replicate(num_simulations, {
  shuffled_flaws <- sample(cables$flaws)
  shuffled_means <- tapply(shuffled_flaws, cables$site, mean)
  #recalc
  null_between_var <- sum(group_sizes * (shuffled_means - overall_mean)^2) / (length(group_means) - 1)
  null_within_var <- sum(tapply(shuffled_flaws, cables$site, function(x) sum((x - mean(x))^2))) / 
                     (nrow(cables) - length(group_means))
  null_between_var / null_within_var
})
#p-value
p_value <- mean(null_stats >= observed_stat)
#results
cat("Test Statistic:", observed_stat, "\n")
cat("p-value:", p_value, "\n")
```

*The observed test statistic is F=2.383, with a corresponding p-value of 0.0971. Since the p-value is greater than the commonly used significance level (α=0.05), we fail to reject the null hypothesis. This result indicates that there is no strong statistical evidence to conclude that the mean number of flaws differs significantly among the three manufacturing facilities (Alderwood, Brentwick, and Coppergrove).*

*While the test suggests no significant differences, it is worth noting that the p-value is relatively close to the threshold of 0.05. This could warrant further investigation, especially if operational differences or other practical considerations suggest potential variations in quality control among the facilities. However, based on the current analysis and data, we do not have sufficient evidence to claim that one facility produces cables with a different level of quality compared to the others.*


## Q2 - Estimation

Let $X_i$ be i.i.d random variables following continuous Uniform($-a,a$) for some unknown positive parameter $a$. Suppose you draw the following sample of $n=15$ observations:

```{r}
data = c(-1.97, -1.07, 0.61, 3.43, -2.51, 3.35, 3.74, 1.35,
         1.08, -3.68, -2.47, -2.72, 1.57, -0.97, 2.27)
```

a. (3pts) Using what you know about estimation, find a reasonable point estimator $\hat a$. Explain why you think this gives a reasonable estimate of the parameter $a$.
   - Note: you do NOT need to find the BEST estimator to earn full points on this part (or later parts), just one that can be argued to be reasonable.
b. (2pts) Apply your estimator to the given sample data. What is your point estimate for $a$?
c. (3pts) Using the MC method shown in class, compute **and interpret** a 95\% confidence interval for $a$. Also, *briefly* (1-2 sentences) explain how well you think this interval would actually perform.
d. (2pts) Just for this part d, suppose the true value of $a$ is actually 3.9. Use MC to approximately estimate the bias of your estimator.

> ***PLEASE WRITE YOUR WORK BELOW:***

> Estimator: $\hat a$ = max(|$X_i$|)

> This estimator is reasonable because for a continuous Uniform(-a,a) distribution, the parameter a defines the largest absolute value any observation can take. Using the maximum absolute value from the sample directly aligns with the definition of a and provides an intuitive estimate of the range of the distribution. Additionally, as the sample size increases, $\hat a$ becomes an increasingly accurate approximation of a. It's a practical and unbiased choice for estimation.

```{r}
# insert code here

#point estimate
a_hat <- max(abs(data))
cat("point estimate for a:", a_hat, "\n")

#MC 95% confidence interval
num_simulations <- 10000
n <- length(data)
simulated_max_values <- replicate(num_simulations, {
  simulated_data <- runif(n, -a_hat, a_hat)
  max(abs(simulated_data))
})
lower_bound <- quantile(simulated_max_values, 0.025)
upper_bound <- quantile(simulated_max_values, 0.975)
mean_simulated_max <- mean(simulated_max_values)
confidence_interval <- c(lower_bound, upper_bound)
cat("95% Confidence Interval for a:", "(", round(confidence_interval[1], 2), ",", round(confidence_interval[2], 2), ")", "\n")
```

> Interpret Confidence Interval: The 95% confidence interval for a is (2.93, 3.73), meaning we are 95% confident that the true value of a, which represents the maximum absolute bound of the uniform distribution’s range, lies within this interval. This interval offers a plausible range for a based on the observed sample data. It was constructed using a Monte Carlo simulation, which approximates the distribution of the test statistic under the assumption of a Uniform(−a,a) distribution. While this method is generally effective, the reliance on extreme values can introduce some variability in small sample sizes (n=15), potentially decreasing the interval’s level of precision in such cases.

```{r}
#new value of a
new_a <- 3.9
#MC simulation -> estimate bias
set.seed(42)
num_simulations <- 10000
n <- 15
simulated_a_hats <- replicate(num_simulations, {
  simulated_data <- runif(n, -new_a, new_a)
  max(abs(simulated_data))
})
#bias calc
mean_simulated_a_hats <- mean(simulated_a_hats)
bias <- mean_simulated_a_hats - new_a
cat("Mean of the estimator:", mean_simulated_a_hats, "\n")
cat("Bias of the estimator:", bias, "\n")
```

> The bias is relatively small compared to the true value of a, suggesting that while $\hat a$ = max(|$X_i$|) may underestimate a slightly, it still provides a reasonable estimate in most cases. For practical purposes, the bias may not significantly affect the usefulness of the estimator, especially for larger sample sizes.

## Q3 - Regression

In R, the built-in dataset `USArrests` contains 1975 estimates of rates (per 100,000 people) of various types of crimes in each US state (source: the World Almanac). Consider the `Assault` and `Murder` columns as X and Y variables:

```{r}
data("USArrests")
x = USArrests$Assault
y = USArrests$Murder
plot(x,y)
```

a. (2pts) Does this look like a reasonable dataset to fit a linear model to? Explain why or why not.
b. (2pts) Regardless of your answer to part a, fit a linear model to this data. Show the summary print out with the full table of coefficients.
c. (2pts) Does this result show a significant relation between x,y? Answer why or why not and give a p-value.
d. (2pts) Give a 95% confidence interval for the true slope.
e. (2pts) Make the standard diagnostic plot(s) for the residuals. Does this fit look reasonable? Explain why or why not.

> ***PLEASE WRITE YOUR WORK BELOW:***

> Based on the scatterplot, it looks reasonable to fit a linear model to this dataset, because there's a visible positive relationship between Assault (x) and Murder (y), with higher assault rates generally corresponding to higher murder rates. The spread of points appears fairly consistent across the range of Assault values, which supports the assumption of homoscedasticity, an important condition for linear modeling. While a few points at higher values might be seen as potential outliers, they don't seem to deviate drastically from the overall trend. Thus, a linear model seems fit for looking into the relationship between these variables.

```{r}
# insert code here
model <- lm(y ~ x)
plot(x,y)
abline(model, col = "blue", lwd = 2)
summary(model)
```

> This result shows a significant relationship between x (Assault) and y (Murder). The p-value for the coefficient of x is 2.6 x 10^-12, which is far below the assumed significance level of 0.05. This low p-value indicates strong evidence against the null hypothesis, suggesting that the rate of assault has a statistically significant positive association with the rate of murder. Therefore, we conclude that there is a significant relationship between assault rates and murder rates in this dataset.

```{r}
confint(model, level = 0.95)["x", ]
```

> This interval suggests that for each additional 100,000 assaults, the murder rate is expected to increase by approximately 0.0328 to 0.0510 per 100,000 people, with 95% confidence. Since zero is not within the interval, this reinforces that the relationship between assault and murder rates is statistically significant.

```{r}
plot(model)
```

*Overall, the diagnostic plots indicate that the linear model is a reasonably good fit for the data, despite some minor caveats. The Residuals vs. Fitted plot shows no strong pattern, supporting the linearity assumption, though slight non-linearity is visible at higher fitted values. The Q-Q plot confirms that residuals are approximately normally distributed, which supports reliable inference despite minor tail deviations. The Scale-Location plot shows consistent variance across fitted values, satisfying the homoscedasticity assumption. Finally, the Residuals vs. Leverage plot reveals no highly influential points, suggesting that no single observation overly affects the model. These diagnostics collectively confirm that the model assumptions are met, making it suitable for analyzing the relationship between assault and murder rates.*
